---
title: "Random forest feature-ranking stability analysis for rbims"
author: "rbims authors"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
params:
  repo_dir: "."
  kegg_dir: "KEGG_hidro"
  metadata_file: "rbims_hydrocarbon-main/Metadata/metadata_SIPH.xlsx"
  output_dir: "reviewer4_rf_stability_results"
  group_col: "Depth"
  feature_col: "KO"
  min_presence: 3
  score_min: 1
  n_iter: 200
  sample_fraction: 0.80
  top_n: 20
  ntree: 1000
  seed: 123
---

# Purpose

This R Markdown document addresses Reviewer 1, comment 4, regarding the random forest component in `rbims`:

> Random forest is used for feature ranking, but no evaluation (e.g., cross-validation, performance metrics, or stability) is provided. The authors should clarify its role and provide validation or justification.

The goal of this analysis is **not** to validate random forest as a predictive or diagnostic model. In `rbims`, random forest is used as an **exploratory feature-ranking tool** that complements ALDEx2-derived compositional effect sizes. Therefore, the most appropriate validation for the current use case is a **feature-ranking stability analysis**.

This document repeats the random forest ranking across resampled datasets and estimates how frequently each KO appears among the top-ranked features.

# Load packages

```{r setup, message=FALSE, warning=FALSE}
required_pkgs <- c(
  "dplyr", "tidyr", "tibble", "ggplot2", "readxl", "readr", "forcats", "purrr", "randomForest"
)

missing_pkgs <- required_pkgs[!vapply(required_pkgs, requireNamespace, logical(1), quietly = TRUE)]
if (length(missing_pkgs) > 0) {
  install.packages(missing_pkgs)
}

suppressPackageStartupMessages({
  library(dplyr)
  library(tidyr)
  library(tibble)
  library(ggplot2)
  library(readxl)
  library(readr)
  library(forcats)
  library(purrr)
  library(randomForest)
  library(rbims)
})

set.seed(params$seed)

repo_dir <- normalizePath(params$repo_dir, mustWork = FALSE)
outdir <- file.path(repo_dir, params$output_dir)
dir.create(outdir, showWarnings = FALSE, recursive = TRUE)
```

# Load rbims input data

This section follows the structure used in the `rbims_hydrocarbon` analysis repository. The expected inputs are:

- KEGG/KofamScan annotations for the 42 MAGs.
- A metadata file containing `Bin_name` and the grouping variable, here `Depth`.

```{r load-data, message=FALSE, warning=FALSE}
# Locate KEGG annotation directory.
kegg_dir_candidates <- c(
  file.path(repo_dir, params$kegg_dir),
  file.path(repo_dir, "KEGG_hidro"),
  file.path(repo_dir, "Functional annotation", "KEGG_hidro"),
  file.path(repo_dir, "Functional_annotation", "KEGG_hidro")
)

kegg_dir <- kegg_dir_candidates[dir.exists(kegg_dir_candidates)][1]
if (is.na(kegg_dir)) {
  stop(
    "Could not find the KEGG annotation directory. Update params$repo_dir and params$kegg_dir in the YAML header."
  )
}

# Locate metadata file.
metadata_candidates <- c(
  file.path(repo_dir, params$metadata_file),
  file.path(repo_dir, "metadata_SIPH.xlsx"),
  file.path(repo_dir, "Metadata", "metadata_SIPH.xlsx"),
  file.path(repo_dir, "metadata", "metadata_SIPH.xlsx")
)

metadata_file <- metadata_candidates[file.exists(metadata_candidates)][1]
if (is.na(metadata_file)) {
  stop(
    "Could not find metadata_SIPH.xlsx. Update params$repo_dir and params$metadata_file in the YAML header."
  )
}

# Read annotations and metadata.
ko_table_raw <- read_ko(data_kofam = kegg_dir)
metadata_raw <- readxl::read_excel(metadata_file) %>% as_tibble()

# Manual MAG name map used in the rbims_hydrocarbon README.
# Names on the left are the raw MAG identifiers in annotation outputs;
# values on the right are the clean MAG names used for plotting and metadata matching.
manual_mag_name_map <- c(
  "5mSIPHEX2_16" = "g_Flavobacterium_5m_16",
  "5mSIPHEX1_26" = "g_Flavobacterium_5m_26",
  "5mSIPHEX1_15" = "g_Henriciella_5m_15",
  "5mSIPHEX1_32" = "g_Hyphomonas_5m_32",
  "5mSIPHEX1_33" = "g_Hyphomonas_5m_33",
  "5mSIPHEX2_10" = "g_Celeribacter_5m_10",
  "5mSIPHEX1_0" = "g_Celeribacter_5m_0",
  "5mSIPHEX1_1" = "s_Planktomarina_temperata_5m_1",
  "5mSIPHEX1_13" = "s_Lentibacter_algarum_5m_13",
  "5mSIPHEX2_7" = "s_Lentibacter_algarum_5m_7",
  "5mSIPHEX2_25" = "g_Tateyamaria_5m_25",
  "5mSIPHEX1_8" = "g_Tateyamaria_5m_8",
  "5mSIPHEX1_2" = "o_Pseudomonadales_5m_2",
  "5mSIPHEX2_5" = "s_Thalassolituus_oleivorans_5m_5",
  "5mSIPHEX1_19" = "s_Thalassolituus_oleivorans_5m_19",
  "5mSIPHEX2_3" = "g_Pseudophaeobacter_5m_3",
  "5mSIPHEX1_37" = "g_Pseudophaeobacter_5m_37",
  "700mSIPHEX1_8" = "g_Pseudophaeobacter_700m_8",
  "700mSIPHEX2_13" = "g_Pseudophaeobacter_700m_13",
  "5mSIPHEX1_9" = "g_Glaciecola_5m_9",
  "700mSIPHEX2_16" = "g_Glaciecola_700m_16",
  "700mSIPHEX1_18" = "g_Glaciecola_700m_18",
  "5mSIPHEX1_11" = "s_Alcanivorax_jadensis_5m_11",
  "700mSIPHEX1_20" = "g_Alcanivorax_700m_20",
  "5mSIPHEX1_25" = "g_Alcanivorax_5m_25",
  "5mSIPHEX2_18" = "g_Alcanivorax_5m_18",
  "5mSIPHEX2_14" = "s_Marinobacter_salarius_5m_14",
  "700mSIPHEX1_3" = "s_Marinobacter_salarius_700m_3",
  "5mSIPHEX1_10" = "s_Marinobacter_salarius_5m_10",
  "700mSIPHEX2_24" = "s_Marinobacter_salarius_700m_24",
  "5mSIPHEX1_18" = "g_Oleibacter_5m_18",
  "700mSIPHEX2_21" = "g_Oleibacter_700m_21",
  "700mSIPHEX1_15" = "g_Oleibacter_700m_15",
  "700mSIPHEX1_17" = "g_Olleya_700m_17",
  "700mSIPHEX2_14" = "g_Olleya_700m_14",
  "700mSIPHEX2_23" = "g_Dokdonia_700m_23",
  "700mSIPHEX1_12" = "g_Dokdonia_700m_12",
  "700mSIPHEX2_9" = "g_Paracoccus_700m_9",
  "700mSIPHEX1_1" = "g_Paracoccus_700m_1",
  "700mSIPHEX1_0" = "g_Sulfitobacter_700m_0",
  "700mSIPHEX2_22" = "g_Alteromonas_700m_22",
  "700mSIPHEX1_2" = "g_Alteromonas_700m_2"
)

rename_mag_columns <- function(x, name_map) {
  current <- names(x)
  hits <- intersect(names(name_map), current)
  if (length(hits) > 0) {
    names(x)[match(hits, current)] <- unname(name_map[hits])
  }
  x
}

rename_mag_values <- function(x, name_map) {
  dplyr::recode(as.character(x), !!!as.list(name_map), .default = as.character(x))
}

stopifnot("Bin_name" %in% names(metadata_raw))
stopifnot(params$group_col %in% names(metadata_raw))

# Helper: convert raw MAG identifiers to clean names when the metadata provides Clean_name.
# This mirrors the rbims_hydrocarbon README, where Bin_name is replaced by Clean_name
# before running mapping_ko().
metadata_use <- metadata_raw %>%
  mutate(
    Bin_name_original = Bin_name,
    Bin_name = if ("Clean_name" %in% names(.)) {
      ifelse(!is.na(Clean_name) & Clean_name != "", Clean_name, Bin_name)
    } else {
      rename_mag_values(Bin_name, manual_mag_name_map)
    },
    Bin_name = rename_mag_values(Bin_name, manual_mag_name_map)
  ) %>%
  select(Bin_name, Bin_name_original, all_of(params$group_col), everything()) %>%
  filter(!is.na(Bin_name), !is.na(.data[[params$group_col]])) %>%
  distinct(Bin_name, .keep_all = TRUE)

# Harmonize annotation names BEFORE mapping_ko().
# read_ko() returns one row per annotation hit and usually contains Bin_name.
# If Bin_name exists, replace raw IDs with Clean_name and then map KOs.
if ("Bin_name" %in% names(ko_table_raw) && "Clean_name" %in% names(metadata_raw)) {
  ko_table_for_mapping <- ko_table_raw %>%
    left_join(
      metadata_raw %>% select(Bin_name, Clean_name),
      by = "Bin_name"
    ) %>%
    mutate(Bin_name = ifelse(!is.na(Clean_name) & Clean_name != "", Clean_name, Bin_name)) %>%
    mutate(Bin_name = rename_mag_values(Bin_name, manual_mag_name_map)) %>%
    select(-Clean_name)
} else if ("Bin_name" %in% names(ko_table_raw)) {
  ko_table_for_mapping <- ko_table_raw %>%
    mutate(Bin_name = rename_mag_values(Bin_name, manual_mag_name_map))
} else {
  ko_table_for_mapping <- rename_mag_columns(ko_table_raw, manual_mag_name_map)
}

# Convert the read_ko() output into the rbims wide KO table used in the vignette.
# If the table is already wide and mapping_ko() is not needed, this will be handled below.
ko_table_mapped <- tryCatch(
  mapping_ko(ko_table_for_mapping),
  error = function(e) {
    message("mapping_ko() failed or was not required. Using read_ko() output directly. Original error: ", e$message)
    ko_table_for_mapping
  }
)

ko_table <- ko_table_mapped

# If mapping_ko() produced raw MAG column names, rename them to clean names.
# First use Clean_name from metadata when available, then apply the manual map from the README.
if ("Clean_name" %in% names(metadata_raw)) {
  rename_vector <- metadata_raw %>%
    filter(!is.na(Clean_name), Clean_name != "") %>%
    select(Bin_name, Clean_name) %>%
    distinct() %>%
    { stats::setNames(.$Clean_name, .$Bin_name) }
  ko_table <- rename_mag_columns(ko_table, rename_vector)
}
ko_table <- rename_mag_columns(ko_table, manual_mag_name_map)

stopifnot(params$feature_col %in% names(ko_table))

# Numeric MAG columns in the rbims table should match metadata_use$Bin_name.
all_numeric_cols <- ko_table %>%
  select(where(is.numeric)) %>%
  names()

mag_cols <- intersect(all_numeric_cols, metadata_use$Bin_name)

# Diagnostics to help troubleshoot remaining name mismatches.
if (length(mag_cols) < 4) {
  message("Numeric columns detected in ko_table:")
  print(utils::head(all_numeric_cols, 20))
  message("metadata_use$Bin_name examples:")
  print(utils::head(metadata_use$Bin_name, 20))
  message("metadata raw Bin_name examples:")
  print(utils::head(metadata_use$Bin_name_original, 20))
  stop(
    "Fewer than four MAG columns matched metadata$Bin_name after Clean_name harmonization. Check whether Clean_name contains the MAG names used as annotation-table columns."
  )
}

metadata_use <- metadata_use %>%
  filter(Bin_name %in% mag_cols) %>%
  select(Bin_name, all_of(params$group_col)) %>%
  mutate(!!params$group_col := factor(.data[[params$group_col]])) %>%
  arrange(match(Bin_name, mag_cols))

stopifnot(nlevels(metadata_use[[params$group_col]]) == 2)

# Keep only metadata-matched MAG columns and non-MAG annotation columns.
# This prevents unmatched numeric columns from entering downstream analyses.
annotation_cols <- setdiff(names(ko_table), all_numeric_cols)
ko_table <- ko_table %>%
  select(all_of(annotation_cols), all_of(mag_cols))

cat("Matched MAGs:", length(mag_cols), "
")
print(table(metadata_use[[params$group_col]]))
cat("First matched MAG columns:
")
print(utils::head(mag_cols, 10))
```

# Run rbims discriminant analysis

This is the same rbims-level analysis used in the manuscript. The random forest stability analysis below evaluates the consistency of the feature-ranking component.

```{r rbims-discriminant, message=FALSE, warning=FALSE}
tibble_disc <- get_subset_discriminant(
  tibble_rbims = ko_table,
  metadata = metadata_use,
  analysis = "KEGG",
  group_col = params$group_col,
  feature_col = params$feature_col,
  min_presence = params$min_presence,
  score_min = params$score_min
)

disc_obj <- attr(tibble_disc, "rbims_disc")

if (is.null(disc_obj)) {
  stop("get_subset_discriminant() did not return the rbims_disc attribute. Check rbims version.")
}

# Save rbims discriminant output.
readr::write_csv(tibble_disc, file.path(outdir, "rbims_discriminant_features.csv"))

if (!is.null(disc_obj$consensus)) {
  readr::write_csv(as_tibble(disc_obj$consensus), file.path(outdir, "rbims_discriminant_consensus.csv"))
}

tibble_disc %>% head(10)
```

# Prepare matrix for random forest stability analysis

Random forest expects samples in rows and features in columns. The code below aggregates duplicated KO rows, filters rare KOs, removes zero-variance features, and aligns the matrix with the metadata.

```{r prepare-matrix}
prepare_rf_matrix <- function(tibble_rbims,
                              metadata,
                              feature_col = "KO",
                              group_col = "Depth",
                              min_presence = 3) {
  mag_cols <- tibble_rbims %>%
    select(where(is.numeric)) %>%
    names()
  mag_cols <- intersect(mag_cols, metadata$Bin_name)

  feature_tbl <- tibble_rbims %>%
    select(all_of(feature_col), all_of(mag_cols)) %>%
    filter(!is.na(.data[[feature_col]])) %>%
    group_by(.data[[feature_col]]) %>%
    summarise(across(all_of(mag_cols), ~ sum(.x, na.rm = TRUE)), .groups = "drop")

  feature_ids <- feature_tbl[[feature_col]]
  x_feature_by_mag <- feature_tbl %>%
    select(all_of(mag_cols)) %>%
    as.matrix()
  rownames(x_feature_by_mag) <- feature_ids

  x <- t(x_feature_by_mag)

  metadata_aligned <- metadata %>%
    filter(Bin_name %in% rownames(x)) %>%
    arrange(match(Bin_name, rownames(x)))

  x <- x[metadata_aligned$Bin_name, , drop = FALSE]
  y <- factor(metadata_aligned[[group_col]])

  keep_presence <- colSums(x > 0, na.rm = TRUE) >= min_presence
  x <- x[, keep_presence, drop = FALSE]

  keep_var <- apply(x, 2, function(z) stats::sd(z, na.rm = TRUE) > 0)
  x <- x[, keep_var, drop = FALSE]

  list(x = x, y = y, metadata = metadata_aligned)
}

rf_input <- prepare_rf_matrix(
  tibble_rbims = ko_table,
  metadata = metadata_use,
  feature_col = params$feature_col,
  group_col = params$group_col,
  min_presence = params$min_presence
)

cat("Samples:", nrow(rf_input$x), "\n")
cat("Features after filtering:", ncol(rf_input$x), "\n")
print(table(rf_input$y))
```

# Random forest feature-ranking stability

For each iteration, the code:

1. Performs stratified subsampling within each group.
2. Trains a random forest classifier on the resampled MAGs.
3. Extracts feature importance scores.
4. Ranks KOs by importance.
5. Records whether each KO appears among the top `r params$top_n` features.

The final stability metric is the frequency with which each KO appears among the top-ranked features across iterations.

```{r stability-function}
run_rf_stability <- function(x,
                             y,
                             n_iter = 200,
                             sample_fraction = 0.8,
                             top_n = 20,
                             ntree = 1000,
                             seed = 123) {
  stopifnot(nrow(x) == length(y))
  y <- factor(y)
  if (nlevels(y) != 2) stop("This stability function expects exactly two groups.")

  set.seed(seed)

  iter_results <- vector("list", n_iter)
  top_sets <- vector("list", n_iter)

  for (i in seq_len(n_iter)) {
    sampled_ids <- unlist(lapply(split(seq_along(y), y), function(idx) {
      n_take <- max(2, floor(length(idx) * sample_fraction))
      n_take <- min(n_take, length(idx))
      sample(idx, size = n_take, replace = FALSE)
    }))

    x_i <- x[sampled_ids, , drop = FALSE]
    y_i <- droplevels(y[sampled_ids])

    rf_fit <- randomForest::randomForest(
      x = x_i,
      y = y_i,
      ntree = ntree,
      importance = TRUE
    )

    imp <- randomForest::importance(rf_fit, type = 2)
    if (is.matrix(imp)) {
      if ("MeanDecreaseGini" %in% colnames(imp)) {
        imp <- imp[, "MeanDecreaseGini"]
      } else {
        imp <- imp[, 1]
      }
    }

    imp_tbl <- tibble(
      iteration = i,
      feature = names(imp),
      importance = as.numeric(imp)
    ) %>%
      arrange(desc(importance), feature) %>%
      mutate(
        rank = row_number(),
        in_top_n = rank <= top_n
      )

    iter_results[[i]] <- imp_tbl
    top_sets[[i]] <- imp_tbl %>%
      filter(in_top_n) %>%
      pull(feature)
  }

  all_iter <- bind_rows(iter_results)

  stability_tbl <- all_iter %>%
    group_by(feature) %>%
    summarise(
      mean_importance = mean(importance, na.rm = TRUE),
      sd_importance = sd(importance, na.rm = TRUE),
      mean_rank = mean(rank, na.rm = TRUE),
      median_rank = median(rank, na.rm = TRUE),
      top_n_frequency = mean(in_top_n, na.rm = TRUE),
      n_iterations = n_distinct(iteration),
      .groups = "drop"
    ) %>%
    arrange(desc(top_n_frequency), mean_rank, desc(mean_importance))

  # Pairwise Jaccard similarity among top-N feature sets.
  if (length(top_sets) >= 2) {
    pair_idx <- utils::combn(seq_along(top_sets), 2)
    jaccard <- apply(pair_idx, 2, function(z) {
      a <- top_sets[[z[1]]]
      b <- top_sets[[z[2]]]
      length(intersect(a, b)) / length(union(a, b))
    })
  } else {
    jaccard <- NA_real_
  }

  list(
    stability = stability_tbl,
    iterations = all_iter,
    jaccard = tibble(
      mean_jaccard_top_n = mean(jaccard, na.rm = TRUE),
      sd_jaccard_top_n = sd(jaccard, na.rm = TRUE),
      min_jaccard_top_n = min(jaccard, na.rm = TRUE),
      max_jaccard_top_n = max(jaccard, na.rm = TRUE)
    )
  )
}
```

```{r run-stability, message=FALSE, warning=FALSE}
rf_stability <- run_rf_stability(
  x = rf_input$x,
  y = rf_input$y,
  n_iter = params$n_iter,
  sample_fraction = params$sample_fraction,
  top_n = params$top_n,
  ntree = params$ntree,
  seed = params$seed
)

stability_tbl <- rf_stability$stability
jaccard_tbl <- rf_stability$jaccard

readr::write_csv(stability_tbl, file.path(outdir, "rf_feature_ranking_stability.csv"))
readr::write_csv(rf_stability$iterations, file.path(outdir, "rf_feature_ranking_all_iterations.csv"))
readr::write_csv(jaccard_tbl, file.path(outdir, "rf_topN_jaccard_summary.csv"))

stability_tbl %>% head(20)
jaccard_tbl
```

# Visualize feature-ranking stability

```{r plot-stability, fig.width=8, fig.height=6}
plot_top_n <- 30

stability_plot <- stability_tbl %>%
  slice_max(order_by = top_n_frequency, n = plot_top_n, with_ties = FALSE) %>%
  mutate(feature = forcats::fct_reorder(feature, top_n_frequency)) %>%
  ggplot(aes(x = top_n_frequency, y = feature)) +
  geom_col() +
  labs(
    x = paste0("Frequency in top ", params$top_n, " across resampling iterations"),
    y = params$feature_col,
    title = "Random forest feature-ranking stability"
  ) +
  theme_bw(base_size = 12)

stability_plot

ggsave(
  filename = file.path(outdir, "rf_feature_ranking_stability_top_features.png"),
  plot = stability_plot,
  width = 8,
  height = 6,
  dpi = 300
)
```

# Join random forest stability with rbims discriminant output

This table helps interpret stable random forest features together with ALDEx2 effect sizes and the rbims consensus output.

```{r join-consensus}
consensus_tbl <- NULL

if (!is.null(disc_obj$consensus)) {
  consensus_tbl <- as_tibble(disc_obj$consensus)

  possible_feature_cols <- c(params$feature_col, "feature", "Feature", "features", "KO")
  consensus_feature_col <- possible_feature_cols[possible_feature_cols %in% names(consensus_tbl)][1]

  if (!is.na(consensus_feature_col)) {
    consensus_tbl <- consensus_tbl %>%
      rename(feature = all_of(consensus_feature_col))

    stability_with_consensus <- stability_tbl %>%
      left_join(consensus_tbl, by = "feature")

    readr::write_csv(
      stability_with_consensus,
      file.path(outdir, "rf_stability_joined_with_rbims_consensus.csv")
    )

    stability_with_consensus %>% head(20)
  } else {
    warning("Could not identify feature column in disc_obj$consensus. Only stability output was saved.")
  }
}
```

# Summary

```{r text-summary, results='asis'}
stable_70 <- stability_tbl %>% filter(top_n_frequency >= 0.70)
stable_50 <- stability_tbl %>% filter(top_n_frequency >= 0.50)

top_features <- stability_tbl %>%
  slice_head(n = 5) %>%
  pull(feature) %>%
  paste(collapse = ", ")

cat("In this analysis, we repeated the random forest feature-ranking procedure ",
    params$n_iter,
    " times using stratified resampling of ",
    params$sample_fraction * 100,
    "% of MAGs within each depth group. For each iteration, we recorded the top ",
    params$top_n,
    " ranked KOs based on MeanDecreaseGini importance. ",
    "The top-ranked features included ", top_features, ". ",
    nrow(stable_70),
    " KOs appeared in the top ", params$top_n,
    " in at least 70% of iterations, and ",
    nrow(stable_50),
    " KOs appeared in the top ", params$top_n,
    " in at least 50% of iterations. ",
    "The mean pairwise Jaccard similarity among top-ranked feature sets was ",
    round(jaccard_tbl$mean_jaccard_top_n, 3),
    ". These results support the use of random forest in rbims as an exploratory feature-prioritization tool rather than as a validated predictive classifier.\n",
    sep = "")
```


# Session information

```{r session-info}
sessionInfo()
```
