---
title: "rbims hydrocarbon case study workflow"
subtitle: "Women in Bioinformatics 2025 revision materials"
author: "Karla P. Lopez-Martinez, Mirna Vazquez-Rosas-Landa, and collaborators"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
    theme: readable
    df_print: paged
params:
  # Set this to the root of the rbims_hydrocarbon repository.
  repo_dir: "."

  # Input folders/files, relative to repo_dir unless absolute paths are provided.
  kegg_dir: "KEGG_hidro"
  interpro_dir: "Interpro_Hidro"
  metadata_file: "rbims_hydrocarbon-main/Metadata/metadata_SIPH.xlsx"
  metadata_renamed_file: "rbims_hydrocarbon-main/metadata_SIPH_renamed.xlsx"
  metadata_5m_file: "rbims_hydrocarbon-main/Metadata/5m_renamed.xlsx"
  metadata_700m_file: "rbims_hydrocarbon-main/Metadata/700m_renamed.xlsx"

  # Output directory, relative to repo_dir unless absolute.
  output_dir: "results/frontiers_women_in_bioinformatics"

  # Optional: path to the reviewer-requested RF stability Rmd.
  rf_stability_rmd: "analysis/rbims_RF_stability_reviewer4.Rmd"

  # Analysis settings.
  min_presence: 3
  score_min: 1
  rf_iterations: 200
  rf_sample_fraction: 0.8
  rf_top_n: 20
  rf_seed: 202501
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center",
  dpi = 300
)
```

# Purpose

This document reproduces the rbims hydrocarbon case study and organizes the additional analyses added during revision. It replaces the previous one-off script with a parameterized R Markdown workflow that can be rendered from the GitHub repository.

The workflow:

1. imports KEGG and InterPro/Pfam annotation outputs;
2. harmonizes MAG names using the metadata table;
3. explores hydrocarbon degradation pathways using rbims;
4. generates the main manuscript figures;
5. runs discriminant analysis and pathway-level directional bias testing;
6. points to the reviewer-requested random forest feature-ranking stability analysis.

# Reproducibility notes

- Run this document from the root of the `rbims_hydrocarbon` repository or edit the YAML parameter `repo_dir`.
- All paths are defined in the YAML header.
- The random forest stability analysis can remain as a separate Rmd. If desired, create a symbolic link from this workflow directory to that Rmd so both analyses are easy to find from the repository.

# Setup

```{r libraries}
required_packages <- c(
  "devtools", "tidyverse", "readxl", "readr", "tibble", "forcats",
  "ComplexHeatmap", "randomForest"
)

missing_packages <- required_packages[!vapply(required_packages, requireNamespace, logical(1), quietly = TRUE)]
if (length(missing_packages) > 0) {
  stop(
    "Missing required packages: ", paste(missing_packages, collapse = ", "),
    "\nInstall them before rendering this document. For rbims, use devtools::install_github('mirnavazquez/RbiMs')."
  )
}

suppressPackageStartupMessages({
  library(tidyverse)
  library(readxl)
  library(readr)
  library(tibble)
  library(forcats)
  library(ComplexHeatmap)
  library(randomForest)
  library(rbims)
})
```

```{r paths}
make_path <- function(path, base_dir = params$repo_dir) {
  if (grepl("^(/|[A-Za-z]:)", path)) return(path)
  file.path(base_dir, path)
}

repo_dir <- normalizePath(params$repo_dir, mustWork = FALSE)
output_dir <- make_path(params$output_dir, repo_dir)
dir.create(output_dir, recursive = TRUE, showWarnings = FALSE)

# Helper to save ComplexHeatmap objects. ggplot2::ggsave() does not work for
# ComplexHeatmap Heatmap/AdditiveUnit objects, so we draw them inside a PDF device.
save_complex_heatmap <- function(ht, filename, width = 9.33, height = 5.50) {
  grDevices::pdf(filename, width = width, height = height)
  ComplexHeatmap::draw(ht)
  grDevices::dev.off()
  invisible(filename)
}

# Locate metadata file.
metadata_candidates <- c(
  file.path(repo_dir, params$metadata_file),
  file.path(repo_dir, "metadata_SIPH.xlsx"),
  file.path(repo_dir, "Metadata", "metadata_SIPH.xlsx"),
  file.path(repo_dir, "metadata", "metadata_SIPH.xlsx")
)
metadata_file <- metadata_candidates[file.exists(metadata_candidates)][1]
if (is.na(metadata_file)) {
  stop(
    "Could not find metadata_SIPH.xlsx. Update params$repo_dir and params$metadata_file in the YAML header."
  )
}
kegg_dir <- make_path(params$kegg_dir, repo_dir)
interpro_dir <- make_path(params$interpro_dir, repo_dir)
#metadata_file <- make_path(params$metadata_file, repo_dir)
metadata_renamed_file <- make_path(params$metadata_renamed_file, repo_dir)
metadata_5m_file <- make_path(params$metadata_5m_file, repo_dir)
metadata_700m_file <- make_path(params$metadata_700m_file, repo_dir)
rf_stability_rmd <- make_path(params$rf_stability_rmd, repo_dir)

stopifnot(dir.exists(kegg_dir))
stopifnot(file.exists(metadata_file))
```

# MAG name harmonization

The rbims tables and metadata must use the same MAG names. We use the `Bin_name` to `Clean_name` mapping stored in the metadata table when it is available. For InterPro/Pfam tables, which may contain raw MAG names as column names, we also define a manual fallback name map.

```{r name-map}
# Read the metadata table and standardize the columns used by rbims.
# This chunk is intentionally defensive because different versions of the
# metadata file may contain either raw MAG IDs, clean MAG names, or both.

manual_mag_name_map <- c(
  "5mSIPHEX2_16" = "g_Flavobacterium_5m_16",
  "5mSIPHEX1_26" = "g_Flavobacterium_5m_26",
  "5mSIPHEX1_15" = "g_Henriciella_5m_15",
  "5mSIPHEX1_32" = "g_Hyphomonas_5m_32",
  "5mSIPHEX1_33" = "g_Hyphomonas_5m_33",
  "5mSIPHEX2_10" = "g_Celeribacter_5m_10",
  "5mSIPHEX1_0"  = "g_Celeribacter_5m_0",
  "5mSIPHEX1_1"  = "s_Planktomarina_temperata_5m_1",
  "5mSIPHEX1_13" = "s_Lentibacter_algarum_5m_13",
  "5mSIPHEX2_7"  = "s_Lentibacter_algarum_5m_7",
  "5mSIPHEX2_25" = "g_Tateyamaria_5m_25",
  "5mSIPHEX1_8"  = "g_Tateyamaria_5m_8",
  "5mSIPHEX1_2"  = "o_Pseudomonadales_5m_2",
  "5mSIPHEX2_5"  = "s_Thalassolituus_oleivorans_5m_5",
  "5mSIPHEX1_19" = "s_Thalassolituus_oleivorans_5m_19",
  "5mSIPHEX2_3"  = "g_Pseudophaeobacter_5m_3",
  "5mSIPHEX1_37" = "g_Pseudophaeobacter_5m_37",
  "700mSIPHEX1_8"  = "g_Pseudophaeobacter_700m_8",
  "700mSIPHEX2_13" = "g_Pseudophaeobacter_700m_13",
  "5mSIPHEX1_9"  = "g_Glaciecola_5m_9",
  "700mSIPHEX2_16" = "g_Glaciecola_700m_16",
  "700mSIPHEX1_18" = "g_Glaciecola_700m_18",
  "5mSIPHEX1_11" = "s_Alcanivorax_jadensis_5m_11",
  "700mSIPHEX1_20" = "g_Alcanivorax_700m_20",
  "5mSIPHEX1_25" = "g_Alcanivorax_5m_25",
  "5mSIPHEX2_18" = "g_Alcanivorax_5m_18",
  "5mSIPHEX2_14" = "s_Marinobacter_salarius_5m_14",
  "700mSIPHEX1_3" = "s_Marinobacter_salarius_700m_3",
  "5mSIPHEX1_10" = "s_Marinobacter_salarius_5m_10",
  "700mSIPHEX2_24" = "s_Marinobacter_salarius_700m_24",
  "5mSIPHEX1_18" = "g_Oleibacter_5m_18",
  "700mSIPHEX2_21" = "g_Oleibacter_700m_21",
  "700mSIPHEX1_15" = "g_Oleibacter_700m_15",
  "700mSIPHEX1_17" = "g_Olleya_700m_17",
  "700mSIPHEX2_14" = "g_Olleya_700m_14",
  "700mSIPHEX2_23" = "g_Dokdonia_700m_23",
  "700mSIPHEX1_12" = "g_Dokdonia_700m_12",
  "700mSIPHEX2_9" = "g_Paracoccus_700m_9",
  "700mSIPHEX1_1" = "g_Paracoccus_700m_1",
  "700mSIPHEX1_0" = "g_Sulfitobacter_700m_0",
  "700mSIPHEX2_22" = "g_Alteromonas_700m_22",
  "700mSIPHEX1_2" = "g_Alteromonas_700m_2"
)

rename_mag_columns <- function(x, name_map) {
  current <- names(x)
  hits <- intersect(names(name_map), current)
  if (length(hits) > 0) {
    names(x)[match(hits, current)] <- unname(name_map[hits])
  }
  x
}

rename_mag_values <- function(x, name_map) {
  dplyr::recode(as.character(x), !!!as.list(name_map), .default = as.character(x))
}

find_first_column <- function(x, candidates) {
  nm <- names(x)
  nm_clean <- tolower(gsub("[^a-z0-9]+", "", nm))
  cand_clean <- tolower(gsub("[^a-z0-9]+", "", candidates))
  hit <- match(cand_clean, nm_clean, nomatch = 0)
  hit <- hit[hit > 0]
  if (length(hit) == 0) return(NULL)
  nm[hit[1]]
}

standardize_metadata <- function(metadata, name_map) {
  names(metadata) <- trimws(names(metadata))

  bin_col <- find_first_column(
    metadata,
    c("Bin_name", "bin_name", "bin", "Bin", "MAG", "MAG_name", "mag_name",
      "Genome", "genome", "Genome_name", "Sample", "sample")
  )

  clean_col <- find_first_column(
    metadata,
    c("Clean_name", "clean_name", "CleanName", "MAG_clean", "mag_clean",
      "Clean_MAG", "clean_mag", "Taxonomic_name", "taxonomic_name")
  )

  if (is.null(bin_col)) {
    stop(
      "Could not find a MAG identifier column in metadata. Available columns are: ",
      paste(names(metadata), collapse = ", "),
      call. = FALSE
    )
  }

  if (bin_col != "Bin_name") {
    metadata <- dplyr::rename(metadata, Bin_name = dplyr::all_of(bin_col))
  }

  if (!is.null(clean_col) && clean_col != "Bin_name" && clean_col != "Clean_name") {
    metadata <- dplyr::rename(metadata, Clean_name = dplyr::all_of(clean_col))
  }

  if (!("Clean_name" %in% names(metadata))) {
    metadata <- metadata %>%
      dplyr::mutate(Clean_name = rename_mag_values(Bin_name, name_map))
  }

  metadata %>%
    dplyr::mutate(
      Bin_name = as.character(Bin_name),
      Clean_name = as.character(Clean_name)
    )
}

metadata_raw <- readxl::read_excel(metadata_file) %>%
  standardize_metadata(manual_mag_name_map)

message("Metadata columns detected: ", paste(names(metadata_raw), collapse = ", "))
message("First Bin_name values: ", paste(utils::head(metadata_raw$Bin_name, 5), collapse = ", "))
message("First Clean_name values: ", paste(utils::head(metadata_raw$Clean_name, 5), collapse = ", "))

metadata_name_map <- metadata_raw %>%
  dplyr::filter(!is.na(Bin_name), !is.na(Clean_name)) %>%
  dplyr::distinct(Bin_name, Clean_name)
```

# KEGG hydrocarbon pathway exploration

```{r read-kegg}
ko_hidro_table <- read_ko(data_kofam = kegg_dir)

ko_hidro_table_renamed <- ko_hidro_table %>%
  dplyr::left_join(metadata_name_map, by = "Bin_name") %>%
  dplyr::mutate(Bin_name = dplyr::coalesce(Clean_name, Bin_name)) %>%
  dplyr::select(-Clean_name)

ko_hidro_mapp_renamed <- mapping_ko(ko_hidro_table_renamed)

metadata_renamed <- if (file.exists(metadata_renamed_file)) {
  readxl::read_excel(metadata_renamed_file)
} else {
  metadata_raw %>%
    dplyr::mutate(Bin_name = dplyr::coalesce(Clean_name, Bin_name)) %>%
    dplyr::select(-Clean_name)
}
```

```{r hydrocarbon-subset}
hydrocarbon_pathways <- c("Naphthalene", "Phenanthrene", "Hexadecane")

hydrocarbon_kos <- ko_hidro_mapp_renamed %>%
  tidyr::drop_na(rbims_pathway) %>%
  get_subset_pathway(rbims_pathway, hydrocarbon_pathways) %>%
  dplyr::mutate(
    rbims_pathway = dplyr::case_when(
      rbims_pathway == "Hexadecane" ~ "HDP",
      rbims_pathway == "Phenanthrene" ~ "PDP",
      rbims_pathway == "Naphthalene" ~ "NPP",
      TRUE ~ rbims_pathway
    )
  )

readr::write_csv(hydrocarbon_kos, file.path(output_dir, "hydrocarbon_kos_mapped.csv"))
```

## Figure 3: pathway coverage bubble plots

```{r orders}
order_taxa_5m <- c(
  "g_Henriciella_5m_15", "g_Hyphomonas_5m_32", "g_Hyphomonas_5m_33",
  "g_Celeribacter_5m_10", "g_Celeribacter_5m_0", "s_Planktomarina_temperata_5m_1",
  "s_Lentibacter_algarum_5m_13", "s_Lentibacter_algarum_5m_7", "g_Tateyamaria_5m_25",
  "g_Tateyamaria_5m_8", "g_Pseudophaeobacter_5m_3", "g_Pseudophaeobacter_5m_37",
  "s_Marinobacter_salarius_5m_14", "s_Marinobacter_salarius_5m_10",
  "g_Flavobacterium_5m_16", "g_Flavobacterium_5m_26", "g_Oleibacter_5m_18",
  "s_Alcanivorax_jadensis_5m_11", "o_Pseudomonadales_5m_2",
  "s_Thalassolituus_oleivorans_5m_5", "s_Thalassolituus_oleivorans_5m_19",
  "g_Glaciecola_5m_9", "g_Alcanivorax_5m_25", "g_Alcanivorax_5m_18"
)

order_taxa_700m <- c(
  "g_Pseudophaeobacter_700m_8", "g_Pseudophaeobacter_700m_13", "g_Paracoccus_700m_9",
  "g_Paracoccus_700m_1", "g_Sulfitobacter_700m_0", "s_Marinobacter_salarius_700m_3",
  "s_Marinobacter_salarius_700m_24", "g_Olleya_700m_17", "g_Olleya_700m_14",
  "g_Dokdonia_700m_23", "g_Dokdonia_700m_12", "g_Alcanivorax_700m_20",
  "g_Alteromonas_700m_22", "g_Alteromonas_700m_2", "g_Oleibacter_700m_21",
  "g_Oleibacter_700m_15", "g_Glaciecola_700m_16", "g_Glaciecola_700m_18"
)

metadata_5m_renamed <- if (file.exists(metadata_5m_file)) readxl::read_excel(metadata_5m_file) else metadata_renamed %>% dplyr::filter(Depth %in% c("5m", "5 m", "Depth_5_meters"))
metadata_700m_renamed <- if (file.exists(metadata_700m_file)) readxl::read_excel(metadata_700m_file) else metadata_renamed %>% dplyr::filter(Depth %in% c("700m", "700 m", "Depth_700_meters"))
```

```{r fig3-bubble-plots, fig.width=9.5, fig.height=5.5}
meta_cols <- c("Module", "Module_description", "Pathway", "Pathway_description", "Cycle",
               "Pathway_cycle", "Detail_cycle", "Genes", "rbims_pathway", "rbims_sub_pathway",
               "Gene_description", "Enzyme", "KO")

hydrocarbon_5m <- hydrocarbon_kos %>%
  dplyr::select(dplyr::any_of(c(meta_cols, order_taxa_5m)))

hydrocarbon_700m <- hydrocarbon_kos %>%
  dplyr::select(dplyr::any_of(c(meta_cols, order_taxa_700m)))

pathways_5m <- plot_bubble(
  tibble_ko = hydrocarbon_5m,
  x_axis = Bin_name,
  y_axis = rbims_pathway,
  analysis = "KEGG",
  calc = "Percentage",
  data_experiment = metadata_5m_renamed,
  order_bins = order_taxa_5m,
  color_character = Class,
  text_x = 11,
  text_y = 11,
  range_size = c(1, 15),
  y_labs = FALSE,
  x_labs = FALSE
)

pathways_700m <- plot_bubble(
  tibble_ko = hydrocarbon_700m,
  x_axis = Bin_name,
  y_axis = rbims_pathway,
  analysis = "KEGG",
  calc = "Percentage",
  data_experiment = metadata_700m_renamed,
  order_bins = order_taxa_700m,
  color_character = Class,
  text_x = 11,
  text_y = 11,
  range_size = c(1, 15),
  y_labs = FALSE,
  x_labs = FALSE
)

pathways_5m
pathways_700m

ggsave(file.path(output_dir, "Figure_3A_pathway_coverage_5m.pdf"), pathways_5m, width = 9.33, height = 5.50, units = "in", dpi = 300, limitsize = FALSE)
ggsave(file.path(output_dir, "Figure_3B_pathway_coverage_700m.pdf"), pathways_700m, width = 9.50, height = 5.50, units = "in", dpi = 300, limitsize = FALSE)
```

## Figure 4: gene-level heatmap

```{r fig4-heatmap, fig.width=8.33, fig.height=5.5}
ko_heatmap <- plot_heatmap(
  tibble_ko = hydrocarbon_kos,
  y_axis = Genes,
  data_experiment = metadata_renamed,
  order_y = rbims_pathway,
  order_x = Depth,
  split_y = TRUE,
  analysis = "KEGG",
  calc = "Abundance"
)

ComplexHeatmap::draw(ko_heatmap)
```

```{r save-fig4, eval=TRUE}
save_complex_heatmap(
  ko_heatmap,
  file.path(output_dir, "Figure_4_Heatmap_of_gene_level_abundances.pdf"),
  width = 8.33,
  height = 5.50
)
```

# Discriminant analysis

```{r discriminant-input-checks}
mag_cols <- ko_hidro_mapp_renamed %>%
  dplyr::select(where(is.numeric)) %>%
  names()

metadata_use <- metadata_renamed %>%
  dplyr::select(Bin_name, Depth) %>%
  dplyr::filter(Bin_name %in% mag_cols) %>%
  dplyr::distinct(Bin_name, .keep_all = TRUE)

missing_in_meta <- setdiff(mag_cols, metadata_use$Bin_name)
if (length(missing_in_meta) > 0) {
  stop("MAG columns not found in metadata_use$Bin_name: ", paste(missing_in_meta, collapse = ", "))
}

print(table(metadata_use$Depth))
stopifnot(nlevels(factor(metadata_use$Depth)) == 2)
```

```{r run-discriminant}
tibble_disc <- get_subset_discriminant(
  tibble_rbims = ko_hidro_mapp_renamed,
  metadata = metadata_use,
  analysis = "KEGG",
  group_col = "Depth",
  feature_col = "KO",
  min_presence = params$min_presence,
  score_min = params$score_min
)

disc_obj <- attr(tibble_disc, "rbims_disc")
stopifnot(!is.null(disc_obj), !is.null(disc_obj$consensus))

consensus_top20 <- disc_obj$consensus %>%
  dplyr::arrange(dplyr::desc(score), dplyr::desc(abs(effect)), dplyr::desc(rf_importance)) %>%
  dplyr::slice_head(n = 20)

readr::write_csv(disc_obj$consensus, file.path(output_dir, "discriminant_consensus_all_kos.csv"))
readr::write_csv(consensus_top20, file.path(output_dir, "discriminant_consensus_top20_kos.csv"))

consensus_top20
```

## Figure 5: discriminant output

```{r fig5-discriminant, fig.width=3.33, fig.height=4.5}
p_heat <- plot_disc_heatmap(
  disc_obj = disc_obj,
  metadata = metadata_use,
  group_col = "Depth",
  top_n = 40
)

p_eff <- plot_disc_effect(disc_obj, top_n = 40)

p_heat
p_eff

ggsave(file.path(output_dir, "Figure_5A_Output_of_the_discriminant_analysis.pdf"), p_heat, width = 3.33, height = 4.50, units = "in", dpi = 300)
ggsave(file.path(output_dir, "Figure_5B_Output_of_the_discriminant_analysis.pdf"), p_eff, width = 3.33, height = 4.50, units = "in", dpi = 300)
```

# Pathway-level directional bias
## Figure 6: Pathway directional bias

```{r pathway-directional-bias, fig.width=7.33, fig.height=5.5}
ko_dictionary <- make_ko_dictionary(
  tibble_rbims = ko_hidro_mapp_renamed,
  feature_col = "KO"
)

pathways <- c("Hexadecane", "Naphthalene", "Phenanthrene")

pathway_tbl <- calc_pathway_directional_bias(
  disc_obj = disc_obj,
  ko_dictionary = ko_dictionary,
  pathways = pathways,
  p_null = 0.5,
  p_alternative = "greater",
  ci_two_sided = TRUE
)

readr::write_csv(pathway_tbl, file.path(output_dir, "pathway_directional_bias_results.csv"))
pathway_tbl

p_ci <- plot_pathway_directional_bias(
  pathway_tbl,
  reorder = TRUE,
  show_fdr_label = TRUE
)

p_ci
ggsave(file.path(output_dir, "Figure_6_Pathway_level_directional_bias.pdf"), p_ci, width = 7.33, height = 5.50, units = "in", dpi = 300)
```

```{r pathway-effects-boxplot, fig.width=7, fig.height=4}
cons_annot <- disc_obj$consensus %>%
  dplyr::left_join(ko_dictionary, by = c("feature" = "KO")) %>%
  dplyr::filter(!is.na(rbims_pathway), !is.na(effect), rbims_pathway %in% pathways)

p_box <- ggplot(cons_annot, aes(x = rbims_pathway, y = effect)) +
  geom_boxplot() +
  theme_bw() +
  labs(
    x = "rbims pathway",
    y = "ALDEx2 effect size",
    title = "Distribution of KO-level effects by hydrocarbon pathway"
  )

p_box
ggsave(file.path(output_dir, "pathway_directional_bias_effect_boxplot.pdf"), p_box, width = 7, height = 4, units = "in", dpi = 300)
```

# InterPro/Pfam protein family analysis
## Figure 7: Protein-level heatmap

```{r interpro-pfam, eval=TRUE, fig.width=9.33, fig.height=5.5}
if (!dir.exists(interpro_dir)) {
  warning("InterPro directory not found: ", interpro_dir, ". Skipping Pfam analysis.")
} else {
  interpro_pfam_profile <- read_interpro(
    data_interpro = interpro_dir,
    database = "Pfam",
    profile = TRUE
  ) %>%
    rename_mag_columns(manual_mag_name_map)

  important_pfams <- get_subset_pca(
    tibble_rbims = interpro_pfam_profile,
    cos2_val = 0.96,
    analysis = "Pfam"
  )

  pfam_heatmap <- plot_heatmap(
    important_pfams,
    y_axis = Pfam,
    analysis = "INTERPRO",
    distance = FALSE
  )

  ComplexHeatmap::draw(pfam_heatmap)
  save_complex_heatmap(
    pfam_heatmap,
    file.path(output_dir, "Figure_7_Heatmap_of_enriched_domain_abundances.pdf"),
    width = 9.33,
    height = 5.50
  )
  readr::write_csv(important_pfams, file.path(output_dir, "important_pfams_pca.csv"))
}
```

# Reviewer-requested random forest stability analysis

The manuscript revision includes a stability assessment of the exploratory random forest feature-ranking step. We keep that analysis as a separate Rmd so it can be rendered independently and linked from the repository.

```{r rf-stability-link, eval=FALSE}
if (file.exists(rf_stability_rmd)) {
  message("RF stability Rmd found: ", rf_stability_rmd)
} else {
  warning(
    "RF stability Rmd not found at: ", rf_stability_rmd,
    "\nPlace the reviewer-requested Rmd at this path or update params$rf_stability_rmd."
  )
}
```

If you want a symbolic link inside this workflow output directory, run the following once from the repository root:

```{bash eval=FALSE}
mkdir -p results/frontiers_women_in_bioinformatics
ln -s ../../analysis/rbims_RF_stability_reviewer4.Rmd \
  results/frontiers_women_in_bioinformatics/rbims_RF_stability_reviewer4.Rmd
```

To render the stability analysis from R:

```{r render-rf-stability, eval=FALSE}
rmarkdown::render(
  input = rf_stability_rmd,
  output_dir = output_dir,
  params = list(
    repo_dir = repo_dir,
    kegg_dir = params$kegg_dir,
    metadata_file = params$metadata_file,
    n_iter = params$rf_iterations,
    sample_fraction = params$rf_sample_fraction,
    top_n = params$rf_top_n,
    seed = params$rf_seed
  )
)
```

Expected reviewer-response outputs from the stability analysis:

- `rf_feature_ranking_stability.csv`
- `rf_topN_jaccard_summary.csv`
- `rf_feature_ranking_all_iterations.csv` if not too large
- `rf_feature_ranking_stability_top_features.png`
- rendered `rbims_RF_stability_reviewer4.html` or `.pdf`

# Session information

```{r session-info}
sessionInfo()
```
